A Practical Optional Type System for Clojure

### 2012-10-10 dissertation/experiments.tex
Monad experiment + other changes

% 
In this chapter we look at several example of using Typed Clojure and
how well the current prototype handles them.

@@

\begin{lstlisting}[caption=call-method, label=lst:callmethod]
(defn call-method
  "Calls a private or protected method.

@@

Thankfully this function has up-to-date documention, and from it we can derive an expected type
(listing \ref{lst:callmethodtype}).

@@

@@

>>>>

In this chapter we look at several example of using Typed Clojure and
how well the current prototype handles them. 
We intentially chose examples that could be challenging to type check
or were particularly useful to the everyday Clojure programmer.

@@

\begin{lstlisting}[caption=call-method, label=lst:callmethod]
;; call-method pulled from clojure.contrib.reflect, (c) 2010 Stuart Halloway & Contributors
(defn call-method
  "Calls a private or protected method.

@@

Thankfully this function has up-to-date documentation, and from it we can derive an expected type
(listing \ref{lst:callmethodtype}).

@@

Now we run the type checker, which produces a type error.

\begin{lstlisting}
#<Exception java.lang.Exception: 29: Cannot call instance method java.lang.reflect.AccessibleObject/setAccessible on type (U nil java.lang.reflect.Method)>
\end{lstlisting}

Because Typed Clojure assumes all methods return nilable Objects, the call to \lstinline|getDeclaredMethod|
has return type \lstinline|(U nil java.lang.reflect.Method)|. It is not type safe to call \lstinline|setAccessible|
on this type, so we get a type error.

In this case, Typed Clojure is too conservative: according to its documentation \lstinline|getDeclaredMethod|
never returns \emph{null}. We add this rule with \lstinline|non-nil-return|.

\begin{lstlisting}
(non-nil-return java.lang.Class/getDeclaredMethod :all)
\end{lstlisting}

Running the type checker produces a different type error.

\begin{lstlisting}
#<Exception java.lang.Exception: Type Error, REPL:32 - (U java.lang.Object nil) is not a subtype of: java.lang.Object>
\end{lstlisting}

This concerns passing \lstinline|obj| as the first argument to the \lstinline|invoke| method.
Typed Clojure conservatively defaults method parameter types as non-nullable. 
Therefore the first parameter of \lstinline|invoke| is \lstinline|Object| by default; \lstinline|obj|
is \lstinline|(U java.lang.Object nil)|. Again, this is too conservative as the first argument can
be \emph{null} for static methods, and we use \lstinline|nilable-param| to specify the first argument
of \lstinline|invoke| may be nil, for the arity of two parameters.

\begin{lstlisting}
(nilable-param java.lang.reflect.Method/invoke {2 #{0}})
\end{lstlisting}

Here is the final successfully type checked code.
\begin{verbatim}Is this a reasonable result?\end{verbatim}

\begin{lstlisting}
(non-nil-return java.lang.Class/getDeclaredMethod :all)
(nilable-param java.lang.reflect.Method/invoke {2 #{0}})

(ann call-method [Class Named (IPersistentVector Class) (U nil Object) (U nil Object) * -> (U nil Object)])

;; call-method pulled from clojure.contrib.reflect, (c) 2010 Stuart Halloway & Contributors
(defn call-method
  "Calls a private or protected method.

  The method-name is given a symbol or a keyword (something Named)."
  [^Class klass method-name params obj & args]
  (let [method (doto (.getDeclaredMethod klass 
                                         (name method-name)
                                         (into-array> Class Class params))
                 (.setAccessible true))]
    (.invoke method obj (into-array> Object (U nil Object) args))))
\end{lstlisting}

@@

\section{Monads}

Monads[] are an interesting control structure probably most recognisable from its inclusion
in the statically-typed language Haskell[].

I chose to port a subset of the Clojure library \emph{algo.monads} to Typed Clojure. The library
provides support for several kinds of monads.

This library represents a monad as a hash-map with four keys: \lstinline|:m-bind|, \lstinline|:m-result|,
\lstinline|:m-zero|, and \lstinline|:m-plus|. A valid monad must provide the first two, and the latter
two may optionally be mapped to the keyword \lstinline|::undefined|\footnote{Keywords prefixed with \lstinline|::|
are qualified in the current namespace.}.

\begin{lstlisting}[caption=Untyped definition for the identity monad, label=lst:identitymdef]
(defmonad identity-m
   "Monad describing plain computations. This monad does in fact nothing
    at all. It is useful for testing, for combination with monad
    transformers, and for code that is parameterized with a monad."
  [m-result identity
   m-bind   (fn m-result-id [mv f]
              (f mv))
  ])
\end{lstlisting}

\begin{lstlisting}[caption=Macro expansion of identity monad definition, label=lst:identitymexpand]
(def identity-m
  (let [m-bind :typed.test.monads/undefined
        m-result :typed.test.monads/undefined
        m-zero :typed.test.monads/undefined
        m-plus :typed.test.monads/undefined
        m-result identity
        m-bind (fn m-result-id [mv f] (f mv))]
    {:m-plus m-plus, :m-zero m-zero, :m-bind m-bind, :m-result m-result}))
\end{lstlisting}

A monad is defined using the macro \lstinline|defmonad|. The macro expands to code that binds a var to a hash-map
with the aforementioned keys. For example, listing \ref{lst:identitymdef} expands to 
the code in listing \ref{lst:identitymexpand}.

\begin{lstlisting}[caption=Type for identity monad, label=lst:identitymtype]
(ann identity-m
     '{:m-bind (All [x y]
                 [x [x -> y] -> y])
       :m-result (All [x]
                   [x -> x])
       :m-zero '::undefined
       :m-plus '::undefined})
\end{lstlisting}

Typed Clojure operates directly on fully macro-expanded forms, so it is unsurprising that the type
annotation for \lstinline|identity-m| reflects its macro-expansion (listing \ref{lst:identitymtype}).
The \lstinline|:m-bind| entry is comparable to Haskell's monadic \emph{bind},
and the \lstinline|:m-result| entry is comparable to Haskell's monadic \emph{return}.

\begin{lstlisting}[caption=Maybe monad definition, label=lst:maybemdef]
(defmonad maybe-m
  "Monad describing computations with possible failures. Failure is
  represented by nil, any other value is considered valid. As soon as
  a step returns nil, the whole computation will yield nil as well."
  [m-zero   nil
   m-result (fn m-result-maybe [v] 
              v)
   m-bind   (fn m-bind-maybe [mv f]
              (if (nil? mv) nil (f mv)))
   m-plus   (fn m-plus-maybe [& mvs]
              (first 
                (filter #(not (nil? a))) mvs))
   ])
\end{lstlisting}

Instead of explaining the identity monad further, we show a more interesting monad, the maybe monad.
Its untyped definition is given in listing \ref{lst:maybemdef}.

\begin{lstlisting}[caption=Typed maybe monad definition, label=lst:maybemtyped]
(ann maybe-m
    '{:m-bind (All [x y]
                [(U nil x) [x -> (U nil y)] -> (U nil y)])
      :m-result (All [x]
                  [x -> (U nil x)])
      :m-zero nil
      :m-plus (All [x]
                [(U nil x) * -> (U nil x)])})
(defmonad maybe-m
  "Monad describing computations with possible failures. Failure is
  represented by nil, any other value is considered valid. As soon as
  a step returns nil, the whole computation will yield nil as well."
  [m-zero   nil
   m-result (ann-form 
              (fn m-result-maybe [v] v)
              (All [x] 
                [x -> (U nil x)]))
   m-bind   (ann-form 
              (fn m-bind-maybe [mv f]
                (if (nil? mv) nil (f mv)))
              (All [x y]
                [(U nil x) [x -> x] -> (U nil x)]))
   m-plus   (ann-form 
              (fn m-plus-maybe [& mvs]
                (first 
                  (filter
                    (ann-form 
                      #(not (nil? %))
                      [(U nil x) -> boolean :filters {:then (is x 0)}])
                    mvs)))
              (All [x]
                [(U nil x) * -> (U nil x)]))
   ])
\end{lstlisting}

A significant difference between typing Haskell style monads in Haskell and typing monads in \emph{algo.monads} with Typed Clojure
is that monads in Haskell are represented by type classes, while Typed Clojure does not support higher-kinded
constructs.

\section{Conduit}

### 2012-10-10 dissertation/intro.tex
Monad experiment + other changes

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. This idea not new,
but recent attempts are noteworthy for their successful use of bidirectional checking.
Instead of inferring types the algorithm relies on programmer annotations.

Clojure is a dynamically typed language running on the JVM. It emphasises
functional programming with immutable data structures
and provides direct interoperability with JVM languages.

Recently languages have been created to or been modified to support aspects
of static and dynamic typing.
Dart is fundamentally dynamically typed but offers a form of optional
static typing. Haskell recently supports the ability to delay static type errors 
to runtime, removing the need for the program to type check before running.
Scala can emulate some features of dynamic languages with Dynamic Scala.

@@

% refinements check invariants of programs
% while ordinary static types check whether programs
% are basically meaningful - Rowan + Frank P
% Page 1 Rowan's PhD
An ordinary static type system is used to check whether programs are basically
meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
static type system relies on programs being well-typed for runtime semantics to be defined.
This means such programs must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
are not defined by the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a static type system that supports exact one type.

@@

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. Runtime contracts are a popular alternative.
Clojure takes this approach, providing easy syntax for defining pre-
and post-conditions \footnote{Inspired by Eiffel}.

@@

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Sam's dissertation TODO}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}
\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.

@@

In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
\lstinline|All| introduces a set of type variables to the body of a type,
here \lstinline|x| is used to define a relationship between the input type and return type.

>>>>

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. This idea not new,
but recent attempts are noteworthy for their successful use of bidirectional checking.
Instead of always attempting to infer types the algorithm relies on programmer annotations
appearing in some natural places, such as giving the type of each top-level function.

The Clojure programming language is a dynamically typed language running on the Java Virtual Machine. 
It emphasises functional programming with immutable data structures
and provides direct interoperability with languages on the Java Virtual Machine.
Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine,
ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
We refer to Clojure the language as \emph{Clojure}, and the family of Clojure-like languages as
\emph{Clojure dialects}.

Clojure dialects are widely used for several reasons.
Performance is a key feature, with most dialects 
offering access to host-level performance

Recently languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart is fundamentally dynamically typed but offers a form of optional
static typing. Haskell recently supports the ability to delay static type errors 
to runtime, removing the need for the program to type check before running.
Scala can emulate some features of dynamic languages with Dynamic Scala.

@@

% refinements check invariants of programs
% while ordinary static types check whether programs
% are basically meaningful - Rowan + Frank P
% Page 1 Rowan's PhD
\begin{verbatim}Following the terminology of [Pfenning, Reynolds ..]\end{verbatim}
An ordinary static type system is used to check whether programs are basically
meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
static type system has run-time semantics that depends on the types of associated
variables and expressions during type checking.
For example, C, Java and ML have intrinsic types.
This means such programs must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
does not depend on the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exact one type,
\begin{verbatim} .. a view strongly advocated by Harper[..]
and common in the literature on static types (see, eg. Pierce[..])
\end{verbatim}

@@

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. Runtime contracts and unit testing are a popular alternatives.
Clojure adopts these approachs, providing easy syntax for defining pre-
and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
\begin{verbatim} but.. (why a static type system is still desirable\end{verbatim}

@@

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{\emph{Example adapted from Sam's dissertation TODO}}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}
\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.

@@

In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}, 
we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
\lstinline|All| introduces a set of type variables to the body of a type,
here \lstinline|x| is used to define a relationship between the input type and return type.

### 2012-10-10 dissertation/lit-review.tex
Monad experiment + other changes

@@

@@

Typed Racket includes safe interoperability between any combination of typed and untyped 
Racket modules\cite{Tob10}\cite{TF08}. 
Interactions with untyped modules are protected by adding runtime checks based on expected types.
Typed Racket implements a sophisticated blame calculus. It ensures 
error messages correctly \emph{blame} the source of type errors,
which can be difficult to determine in the presence of higher-order functions. %TODO reference

@@

Forsyth\cite{Rey96} was the first language to use intersection types.
Uses of intersections in Forsyth include representing extensible record types
and function overloading.

% CIDRE
% TODO
CIDRE\cite{Dav05} adds \emph{refinement type checking} to SML.

% PADL '12 paper, Typing the Numeric Tower
Typed Racket uses \emph{ordered intersection types} to offer a kind of function overloading.

Java allows intersection types only in type parameter bounds. Instead of supporting intersection
types in the rest of the language, Java support type inheritance by extending Classes and Interfaces.

@@

Designing a static type system for any dynamically typed language requires
careful consideration. Idiomatic style in dynamically typed languages is usually
designed to be powerful yet non-intrusive. An effective type system
for such languages should preserve existing idioms and syntactic abstractions
until time tested idioms are devised for the new typed language \cite{Tob10}.

Typed Racket \cite{Tob10} integrates a selection of ideas to
create a typed sister language to Racket.
Clojure and Racket share a key similarity: they are dynamically typed Lisps.
Consequently, a number of novel Typed Racket features are relevant to the design and implementation
of a type system for Clojure.
The inventions of occurrence typing \cite{TF10}
and a systematic approach to variable-arity polymorphism
\cite{STF09}
are both applicable to this project.

@@

Clojure and the statically typed language Scala \cite{OCD+} 
have some interesting similarities. The redesigned 
of the Scala collection hierarchy
\cite{OS10}
bears similarity to Clojure's encompassing ``seq''
model, revealing an opportunity to test potential typed extensions to the Clojure 
class hierarchy.

@@

\section{Variable-Arity Polymorphism}

Functions with variable-arity are abundant in dynamically typed languages;
it is common to have a large standard library of functions that utilize
variable-arity, and Clojure is no different. 
Support for complex variable-arity functions is sparse in typed languages,
Strickland, Tobin-Hochstadt, and Felleisen claim \cite{STF09}
``no typed language supports them in a systematic and principled manner''.

Strickland et al. invented \cite{STF09} 
a calculus capable of typing non-uniform variable-arity functions, a version of which
is implemented in Typed Racket. A non-uniform variable-arity function is a function that
takes a heterogeneous variable parameter list, which usually has a complex relationship 
with other components of the function.
For example, Clojure's \lstinline|map| function takes a function and one or more sequences,
and returns the result of applying the function argument to each element of the sequences pair-wise.

\begin{lstlisting}[caption=An application of the non-uniform variable-arity function \lstinline|map|, label=lst:map]
(map + [1 2] [2.1 3.2]) 
;=> (3.1 5.2)
\end{lstlisting}

@@

Clojure presents some opportunities to extend Strickland's calculus. It is idiomatic
to utilize hash-maps as ad-hoc objects in Clojure, and it would be useful to track the keys associated
with a hash-map.  The fundamental hash-map manipulation functions \lstinline|assoc| and \lstinline|dissoc|,
which associate and dissociate entries to (immutable) hash-maps, both take a variable number of arguments.

\begin{lstlisting}[caption=Using \lstinline|dissoc| with variable arguments, label=lst:assoc1]
(dissoc {:a 1, :b 2} ; => {:a 1, :b 2}
        :a           ; => {:b 2}
        :b}          ; => {}
;=> {}
\end{lstlisting}

@@

\begin{lstlisting}[caption=Bidirectional checking algorithm with Typed Clojure pseudocode, label=lst:bidir]
(map (fn [a :- Long, b :- Float]
       (+ a b))
     [1 2]
     [2.1 3.2])
;=> (3.1 5.2)
\end{lstlisting}

@@

The Local Type Inference paper \cite{PT00}
describes two implementations of type variables, for bounded
and unbounded type variables. The bounded implementation is presented
as an optional extension  to the unbounded implementation, which preserves all
properties described in the Local Type Inference algorithm.

>>>>

In this chapter we fit the design of Typed Clojure with literature from related fields.
Forward-references are given to link the discussed literature with this project
where there is some relevance or influence.

@@

\begin{verbatim}
  Different levels of Gradual Typing discussed by Siek..
\end{verbatim}

@@

Typed Racket includes safe interoperability between any combination of typed and untyped 
Racket modules\cite{Tob10}\cite{TF08}. 
Interactions with untyped modules are protected by adding runtime checks based on expected types.
Typed Racket implements a sophisticated blame calculus. It ensures 
error messages correctly \emph{blame} the source of type errors,
which can be difficult to determine in the presence of higher-order functions. 
\begin{verbatim}references, Walder etc.\end{verbatim}

@@

Forsythe, a modern ALGOL dialect by Reynolds\cite{Rey96} was the first wide-spread 
programming language to use intersection types.
Uses of intersections in Forsythe include representing extensible record types
and function overloading.

% CIDRE
SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
that features a form of intersection types. It is used for function overloading
and checking invariants involving Refinement types.
\begin{verbatim}
Freeman & Pfenning, Refinement types
\end{verbatim}

Typed Racket, developed by Tobin-Hochstadt and others \cite{Tob10}, uses \emph{ordered intersection types} 
to offer a kind of function overloading.
\begin{verbatim}PADL '12 paper, Typing the Numeric Tower\end{verbatim}

@@

@@

Clojure and the statically typed language Scala \cite{OCD+} 
have some interesting similarities. The redesign
of the Scala collection hierarchy\cite{OS10}
bears similarity to Clojure's encompassing ``seq''
model, revealing an opportunity to test potential typed extensions to the Clojure 
class hierarchy.

@@

\section{Function Types}

There are several different approaches to representing functions in programming languages.

In typed languages like Haskell, functions are as simple as possible, taking a single argument.
A function with multiple arguments is represented by chaining several single-argument function
together. This style is characterised by direct syntactic function currying, where applying
a function to less than its maximum number of arguments results in another function
that takes the remaining arguments.

In untyped languages like Scheme, functions can take any number of arguments. It is an
error to supply less than the minimum number of arguments to a function.
This style features sophisticated support for functions with variable-arity. For example,
functions can dispatch on the number of arguments provided, and supports a \emph{rest} parameter
as its last parameter which can accept any number of arguments.

In this regard, Clojure takes an approach identical to Scheme, and supports all the features
mentioned in the previous paragraph, and none characterised by Haskell-style functions.
For this reason, we ignore the tradeoffs associated between the two approaches 
and move directly to literature applicable to typing Scheme-style functions.

\subsection{Variable-Arity Polymorphism}

Strickland et al. invented a type system supporting variable-arity polymorphism\cite{STF09}
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

For example in Clojure, the function \lstinline|map| takes a function and one or more sequences,
and returns the result of applying the function argument to each element of the sequences pair-wise.

\begin{lstlisting}[caption=An application of the non-uniform variable-arity function \lstinline|map|, label=lst:map]
(map + [1 2] [2.1 3.2]) 
;=> (3.1 5.2)
\end{lstlisting}\footnote{Line comments in Clojure begin with \lstinline|;| and comments to the end of the line. We use \lstinline|;=>| to mean \emph{evaluates to}.}

@@

@@

\begin{lstlisting}[caption=Bidirectional checking algorithm with Typed Clojure pseudocode, label=lst:bidir]
(map (fn [[a :- Long] [b :- Float]]
       (+ a b))
     [1 2]
     [2.1 3.2])
;=> (3.1 5.2)
\end{lstlisting}

@@

Local type inference by Pierce and Turner\cite{PT00}
describe two implementations of type variables, for bounded
and unbounded type variables. The bounded implementation is presented
as an optional extension  to the unbounded implementation, which preserves all
properties described in the Local Type Inference algorithm.

### 2012-10-10 dissertation/paper.tex
Experiments
Monad experiment + other changes

\title{Typed Clojure: A Practical Optional Type System for Clojure}
\author{Ambrose Bonnaire-Sergeant}

@@

\begin{abstract}
  % FIXME long tradition for attempts
  There is a long tradition to create static type systems for dynamically-typed
  languages. Static type checking promises to eliminate many common user errors
  that are otherwise unnoticed or are caught when the program is run.

  Recent attempts have seen several promising and innovative
  type systems built for dynamically typed languages. Typed Racket is able to
  statically type check idioms ..
  We take the lessons learnt from these
  projects and apply them to the dynamically-typed language Clojure in the form of Typed Clojure.
\end{abstract}

@@

>>>>

\title{A Practical Optional Type System for Clojure}
\author{Ambrose Bonnaire-Sergeant}

@@

\begin{abstract}
  % FIXME long tradition for attempts
  Since the first wide-spread programming languages like Fortran, LISP and ALGOL
  there has been a division between languages that perform type checking at compile-time
  and those that don't.
  A recent trend is to combine the advantages of both kinds of languages by adding optional
  static type systems to languages without static type checking,
  which promises to eliminate many common user errors
  that are otherwise unnoticed or are caught when the program is run.

\begin{verbatim}
  Recent attempts have seen several promising and innovative
  type systems built for dynamically typed languages. Typed Racket is able to
  statically type check idioms ..
\end{verbatim}
  We take the lessons learnt from these
  projects and apply them to the dynamically-typed language Clojure in the form of Typed Clojure.
\end{abstract}

@@

\include{experiments}
